---
title: "PFAS and Eicosanoids"
output: html_notebook
---

This notebook will include analysis related to PFAS and Eicosanoids in both PROTECT and LIFECODES

### Import data 

Load some libraries

```{r loading libraries}

library(dplyr)
library(tidyr)
library(ggplot2)
library(naniar)
library(corrplot)
library(patchwork)
library(ggpubr)
library(pheatmap)
library(forcats)

```

Import some data files 

```{r import files}

pfas_eic_combined <- read.csv("pfas_eic_pr_lc_comb.csv") %>% select(-X)
covars_lc_pr <- read.csv("covars_pr_lc_comb.csv")%>% select(-X)
covars_lc <- read.csv("covars_lc.csv")%>% select(-X)
covars_pr <- read.csv("covars_pr.csv")%>% select(-X)
eic_class <- readxl::read_xlsx(path = "eicosanoids_comparisons.xlsx", sheet = "Eicosanoids_LC_PR") %>% 
  select(`PR Name`, Group2)
names(eic_class) <- c("vars", "group")
eic_class$vars <- tolower(eic_class$vars)

# PFAS concentration columns
exposures_all<- grep("^conc_", names(pfas_eic_combined), value = TRUE)
exposures_short <- c("conc_pfoa", "conc_pfos", "conc_pfhxs", "conc_pfna")
exposures_lc <- setdiff(exposures_all, c("conc_pfhpa", "conc_pfosa" ))


covars_list <- c("age", "bmi_prepreg", "as.factor(mat_edu_cat)", "as.factor(ins_cat)", "race4b", "nullip", "as.factor(cohort)")
covars_list0 <- c("age", "bmi_prepreg", "mat_edu_cat", "ins_cat", "race4b", "nullip", "cohort")


covars_list_lc <- c("age", "bmi_prepreg", "as.factor(mat_edu_cat)", "as.factor(ins_cat)", "race4b", "nullip", "year_t1")
pfas_eic_lc <- pfas_eic_combined2 %>% filter(cohort == "LC")


covars_list_pr <- c("age", "bmi_prepreg", "as.factor(mat_edu_cat)", "as.factor(ins_cat)", "nullip", "year_t1")
pfas_eic_pr <- pfas_eic_combined2 %>% filter(cohort == "PR")


#Two eicosanoids ltc4.me and lte4 have 0 as minimim values. This will cause an error for these eicosanoids later on

pfas_eic_combined <- pfas_eic_combined %>%
  mutate(lte4 = ifelse(lte4 == 0, NA, lte4),
         ltc4.me = ifelse(ltc4.me == 0, NA, ltc4.me))

eic_all <- names(pfas_eic_combined[,21:75])
eic_all2 <- setdiff(eic_all, c("dipge2", "dipgj2"))

pfas_eic_combined2 <- pfas_eic_combined %>% 
  inner_join(covars_lc_pr, by = c("cohort", "id")) %>% 
  distinct() %>%
  mutate(race4b = ifelse(race4 == "Hispanic-PR", "Hispanic", race4))

# pfas_eic_combined2 %>% group_by(cohort, id) %>% tally() %>% filter(n >1) 
# 
#Add third visit ga
tvdate_ga <- read.csv("thirdvisitdate.csv") %>% select(-"X")

pfas_eic_combined2 <- pfas_eic_combined2 %>% left_join(tvdate_ga, by = c("id", "cohort"))

```

### Distributions of PFAS by cohorts 

First we will look at the distributions of PFAS concentrations in the study participants by cohort

```{r distribution of PFAS by cohort}

pfas_summary_results <- lapply(exposures_all, function(col) {
  pfas_col_summary(pfas_eic_combined, col)
}) %>%
  bind_rows() %>%
  select(PFAS, everything()) %>%
  arrange(PFAS, cohort)

# Print results
print(pfas_summary_results)

# Optional: Save to CSV
write.csv(pfas_summary_results, "pfas_conc_summary_by_cohort.csv", row.names = FALSE)

```

We have 4 PFAS with detection rate above 50% across both cohorts: PFNA, PFHxS, PFOA, and PFOS

Focus on these 4 PFAS for now

```{r correlation between PFAS}

create_heatplot(pfas_eic_combined, exposures_short, "")
create_heatplot(pfas_eic_combined, exposures_short, "cohort")

```

In LIFECODES, the strongest correlation is between PFOA and PFOS, while in PROTECT, the strongest correlation is between PFOA and PFNA. 

Distributions of PFAS in LIFECODES and PROTECT

```{r distributions of PFAS in LIFECODES and PROTECT}

hist_pfas <- pfas_eic_combined %>% select(id, cohort, all_of(exposures_short)) %>%
  pivot_longer(cols = conc_pfoa:conc_pfna, names_to = "pfas", values_to = "pfas_conc")

hist_pfas %>% ggplot(aes(y = log(pfas_conc), col = cohort)) +
  geom_boxplot() +
  facet_wrap(~ pfas, scales = "free")


```

### Correlation between different eicosanoids

Now we will look at correlation between different eicosanoids 

First, in the overall dataset

```{r correlation between eicosanoids in the overall dataset}

# Ensure annotation is sorted by group
eic_class <- eic_class %>% arrange(group)
annotation_row_label <- data.frame(group = as.factor(eic_class$group))
rownames(annotation_row_label) <- eic_class$vars

cor_eic_func(data = pfas_eic_combined, cohort_variable = "All")

```

Then only in LIFECODES

```{r correlation between eicosanoids in LC}

cor_eic_lc <- cor_eic_func2(data = pfas_eic_combined, cohort_variable = "LC")

```

And then only in PROTECT

```{r correlation between eicosanoids in PR}

cor_eic_pr <- cor_eic_func2(data = pfas_eic_combined, cohort_variable = "PR")

```

### Get the most important correlation [eho > abs(0.5)]

```{r choosing the higher correlation}

cor_to_choose_lc <- which(cor_eic_lc >= 0|cor_eic_lc <=0, arr.ind = TRUE)
cor_to_choose_pr <- which(cor_eic_pr >= 0|cor_eic_pr <=0, arr.ind = TRUE)


# Create a data frame of the results
cor_to_choose_lc2 <- data.frame(
  Variable1 = rownames(cor_eic_lc)[cor_to_choose_lc[, 1]],
  Variable2 = colnames(cor_eic_lc)[cor_to_choose_lc[, 2]],
  Correlation = cor_eic_lc[cor_to_choose_lc]
)

cor_to_choose_pr2 <- data.frame(
  Variable1 = rownames(cor_eic_pr)[cor_to_choose_pr[, 1]],
  Variable2 = colnames(cor_eic_pr)[cor_to_choose_pr[, 2]],
  Correlation = cor_eic_pr[cor_to_choose_pr]
)

cor_eic_lc3 <- cor_eic_func2(data = pfas_eic_combined, cohort_variable = "LC", threshold = 0.5)
cor_eic_pr3 <- cor_eic_func2(data = pfas_eic_combined, cohort_variable = "PR", threshold = 0.5)

#Compare correlation between eicosanoids in PR and LC
cor_to_choose_lc2b <- cor_to_choose_lc2 %>% filter(Variable1 != Variable2)
cor_to_choose_pr2b <- cor_to_choose_pr2 %>% filter(Variable1 != Variable2)

cor_to_choose_both <- cor_to_choose_lc2b %>% 
inner_join(cor_to_choose_pr2b, by = c("Variable1", "Variable2"))

ggplot(cor_to_choose_both, aes(x = Correlation.x, y = Correlation.y)) +
  geom_point() +
  xlab("Correlation coefficients in LC") + ylab("Correlation coefficients in PR")

cor(x = cor_to_choose_both$Correlation.x, 
    y = cor_to_choose_both$Correlation.y, method = "pearson") 

```
Check the distributions of eicosanoids in LC and PR

```{r distributions of eicosanoids in LC and PR}

pfas_eic_combined_long <- pfas_eic_combined %>% select(id, cohort, dipgf2:dipgj2) %>%
  pivot_longer(cols = dipgf2:dipgj2, names_to = "vars", values_to = "eic_value") %>%
  inner_join(eic_class, by = "vars")

eic_hist_fun <- function(group_name){
  p <- pfas_eic_combined_long %>%
    filter(group == group_name) %>%
    ggplot(aes(x = log(eic_value), col = cohort)) +
    geom_density() +
    facet_wrap(~vars, nrow = 3, scales = "free") +
    ggtitle(paste(group_name, "group"))
  return(p)
}

# Now call the function
p1eic <- eic_hist_fun(group_name = "Cyclooxygenase")
p2eic <- eic_hist_fun(group_name = "Cytochrome p450")
p3eic <- eic_hist_fun(group_name = "Lipoxygenase")
p4eic <- eic_hist_fun(group_name = "Parent Compound")

p1eic
p2eic
p3eic
p4eic

```

Generally, eicosanoid levels are higher in PROTECT as compared to LIFECODES

### Correlation between PFAS and eicosanoids 

```{r correlation between PFAS and eicosanoids}

annotation_row_label2 <- data.frame(group = as.factor(eic_class$group))
rownames(annotation_row_label2) <- eic_class$vars

cor_eic_pfas_func(data = pfas_eic_combined, cohort_variable = "PR",
                   pfas_vec = exposures_short)
chkpr <- cor_eic_pfas_func(data = pfas_eic_combined, cohort_variable = "PR",
                   pfas_vec = exposures_short) 

cor_eic_pfas_func(data = pfas_eic_combined, cohort_variable = "LC",
                    pfas_vec = exposures_short)
chklc <- cor_eic_pfas_func(data = pfas_eic_combined, cohort_variable = "LC",
                    pfas_vec = exposures_lc)
```

### Adjusted models for PFAS and Eicosanoids - both cohorts

Now we will model the relationship between eicosanoids and PFAS taking into consideration the covariates 

```{r running adjusted model between PFAS and eicosanoids}

eic_chk1 <- lmerTest::lmer(log(pgj2) ~ log(conc_pfoa) + age + bmi_prepreg + as.factor(mat_edu_cat) + ins_cat + race4 +
                  year_t1 + nullip + (1|cohort), data = pfas_eic_combined2)
summary(eic_chk1)

eic_chk2 <- lmerTest::lmer(log(pgj2) ~ log(conc_pfos) + age + bmi_prepreg + as.factor(mat_edu_cat)  + ins_cat + race4 +
                  year_t1 + nullip + (1|cohort), data = pfas_eic_combined2)
summary(eic_chk2)


eic_chk3 <- glm(log(pgj2) ~ log(conc_pfoa) + age + bmi_prepreg + as.factor(mat_edu_cat)  + ins_cat + race4b +
                  year_t1 + nullip, data = pfas_eic_combined2)
summary(eic_chk3)

eic_chk4 <- glm(log(pgj2) ~ log(conc_pfoa) + age + bmi_prepreg + as.factor(mat_edu_cat)  + ins_cat + race4b +
                  year_t1 + nullip + cohort, data = pfas_eic_combined2)
summary(eic_chk4)

eic_chk5 <- glm(log(pgj2) ~ log(conc_pfoa) + age + bmi_prepreg + as.factor(mat_edu_cat)  + ins_cat + race4b +
                 nullip + cohort, data = pfas_eic_combined2)
summary(eic_chk5) #provides the best fit for now

eic_chk6 <- glm(log(pgj2) ~ log(conc_pfoa) + age + bmi_prepreg + as.factor(mat_edu_cat)  + ins_cat + race4b +
                 nullip , data = pfas_eic_combined2)
summary(eic_chk6)

eic_chk7 <- glm(log(pgj2) ~ log(conc_pfoa) + age + bmi_prepreg + as.factor(mat_edu_cat)  + ins_cat + race4b +
                 nullip + log(conc_pfoa)*cohort, data = pfas_eic_combined2)
summary(eic_chk7)


```

Now call the function

```{r adjusted model between PFAS and eicosanoids - both cohorts}

# Run the function
results_both <- pfas_eic_adjmod(data = pfas_eic_combined2, exposures = exposures_short, 
                                outcomes = eic_all2, covariates = covars_list)


# Now I will do a heatplot of results 

results_both2 <- results_both %>% mutate(vars = Outcome) %>%
  inner_join(eic_class, by = "vars") %>% select(-vars)

names(results_both2) <- c("exposure", "outcome", "estimate", "conf.low", "conf.high", "pval", "group")

plot_adj_result <- function(data, pfas){
data %>%
    filter(exposure == pfas) %>%
    ggplot(aes(x = estimate, xmin = conf.low, xmax = conf.high, y = outcome, col = group)) +
    geom_pointrange() +
    theme(legend.position = "none") +
    geom_vline(xintercept = 0) +
    xlab(paste0("Estimates for ", toupper(sub("conc_", "", pfas)))) +
    facet_grid(group ~ ., scales = "free_y", switch = "y") +  # Facet by group, switch facet labels to y-axis
    theme(strip.placement = "outside", strip.text.y = element_text(angle = 0),
          axis.title.y = element_blank()) }

pfoa_comb <- plot_adj_result(data = results_both2, pfas = "conc_pfoa") + theme(strip.placement = "outside")
pfos_comb <- plot_adj_result(data = results_both2, pfas = "conc_pfos") + theme(axis.text.y = element_blank(), axis.ticks.y = element_blank(), strip.text.y = element_blank())
pfna_comb <- plot_adj_result(data = results_both2, pfas = "conc_pfna") + theme(axis.text.y = element_blank(), axis.ticks.y = element_blank(),  strip.text.y = element_blank())
pfhxs_comb <- plot_adj_result(data = results_both2, pfas = "conc_pfhxs") + theme(axis.text.y = element_blank(), axis.ticks.y = element_blank(),  strip.text.y = element_blank())

# Combine plots using patchwork
combined_plot <- pfoa_comb | pfos_comb | pfna_comb | pfhxs_comb

combined_plot


```

Another way of plotting the results is to just do heatplots

```{r heatplot of esitmates}

results_both2 <- results_both2 %>%
    mutate(significant = ifelse(pval < 0.05, "*", ""))

# Create the heatmap with faceting by group
ggplot(results_both2, aes(x = exposure, y = outcome, fill = estimate)) +
    geom_tile(color = "white") +
    geom_text(aes(label = significant), color = "black", size = 5, vjust = 0.5) +
    scale_fill_gradient2(low = "blue", mid = "white", high = "red", midpoint = 0) +
    theme_minimal() +
    theme(
        axis.text.x = element_text(angle = 45, hjust = 1, size = 8),  # Adjust x-axis text size and angle
        axis.text.y = element_text(size = 6),  # Adjust y-axis text size
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        strip.placement = "outside",
        strip.text.y = element_text(angle = 180),
        legend.position = "bottom"
    ) +
    facet_grid(group ~ ., scales = "free_y", switch = "y") +
    ggtitle("Heatmap of PFAS-Eicosanoid/Parent Compounds Associations")


```

### Adjusted associations for PFAS and Eicosanoids - LC only

```{r adjusted associations between PFAS and Eicosanoids - LC only}

# Run the function for LC
results_lc <- pfas_eic_adjmod(data = pfas_eic_lc, exposures = exposures_lc, 
                                outcomes = eic_all2, covariates = covars_list_lc)

results_lc2 <- results_lc %>% mutate(vars = Outcome) %>%
  inner_join(eic_class, by = "vars") %>% select(-vars)

names(results_lc2) <- c("exposure", "outcome", "estimate", "conf.low", "conf.high", "pval", "group")


pfoa_lc<- plot_adj_result(data = results_lc2, pfas = "conc_pfoa") + theme(strip.placement = "outside") + ggtitle("Associations in LC")
pfos_lc<- plot_adj_result(data = results_lc2, pfas = "conc_pfos") + theme(axis.text.y = element_blank(), axis.ticks.y = element_blank(), strip.text.y = element_blank())
pfna_lc<- plot_adj_result(data = results_lc2, pfas = "conc_pfna") + theme(axis.text.y = element_blank(), axis.ticks.y = element_blank(),  strip.text.y = element_blank())
pfhxs_lc<- plot_adj_result(data = results_lc2, pfas = "conc_pfhxs") + theme(axis.text.y = element_blank(), axis.ticks.y = element_blank(),  strip.text.y = element_blank())
pfua_lc<- plot_adj_result(data = results_lc2, pfas = "conc_pfua") +  theme(axis.text.y = element_blank(), axis.ticks.y = element_blank(),  strip.text.y = element_blank())
pfdea_lc<- plot_adj_result(data = results_lc2, pfas = "conc_pfdea") + theme(axis.text.y = element_blank(), axis.ticks.y = element_blank(),  strip.text.y = element_blank())
mpah_lc<- plot_adj_result(data = results_lc2, pfas = "conc_me.pfosa.acoh") + theme(axis.text.y = element_blank(), axis.ticks.y = element_blank(),  strip.text.y = element_blank())

# combine plots using patchwork
lc_plot <-  pfoa_lc | pfos_lc | pfna_lc | pfhxs_lc | pfua_lc | pfdea_lc | mpah_lc

lc_plot 


```

### Adjusted associations for PFAS and Eicosanoids - PR only

```{r adjusted associations between PFAS and Eicosanoids - PR only}

# Run the function for PR
results_pr <- pfas_eic_adjmod(data = pfas_eic_pr, exposures = exposures_short, 
                                outcomes = eic_all2, covariates = covars_list_pr)

results_pr2 <- results_pr %>% mutate(vars = Outcome) %>%
  inner_join(eic_class, by = "vars") %>% select(-vars)

names(results_pr2) <- c("exposure", "outcome", "estimate", "conf.low", "conf.high", "pval", "group")


pfoa_pr<- plot_adj_result(data = results_pr2, pfas = "conc_pfoa") + theme(strip.placement = "outside") + ggtitle("Associations in PR")
pfos_pr<- plot_adj_result(data = results_pr2, pfas = "conc_pfos") + theme(axis.text.y = element_blank(), axis.ticks.y = element_blank(), strip.text.y = element_blank())
pfna_pr<- plot_adj_result(data = results_pr2, pfas = "conc_pfna") + theme(axis.text.y = element_blank(), axis.ticks.y = element_blank(),  strip.text.y = element_blank())
pfhxs_pr<- plot_adj_result(data = results_pr2, pfas = "conc_pfhxs") + theme(axis.text.y = element_blank(), axis.ticks.y = element_blank(),  strip.text.y = element_blank())

# combine plots using patchwork
pr_plot <- pfoa_pr | pfos_pr | pfna_pr | pfhxs_pr 

pr_plot 


```

Compare results between PR and LC 

```{r compare results between PR and LC}

results_lc_pr2 <- rbind.data.frame((results_pr2 %>% mutate(cohort = "PR")),
                                   (results_lc2 %>% mutate(cohort = "LC")))


plot_adj_result2 <- function(data, pfas, subset = "All", width = 1){
  if(subset == "All"){
  data2 <- data}
  else {
    data2 <- data %>% filter(group == subset) 
  }
  data2 <- data2 %>%
    filter(exposure == pfas) %>%
    ggplot(aes(x = estimate, xmin = conf.low, xmax = conf.high, y = outcome, col = cohort)) +
    geom_pointrange(position = position_dodge(width = 0.5)) +
    theme(legend.position = "none") +
    geom_vline(xintercept = 0) +
    xlab(paste0("Estimates for ", toupper(sub("conc_", "", pfas)))) +
    facet_grid(group ~ ., scales = "free_y", switch = "y") +  # Facet by group, switch facet labels to y-axis
    theme(strip.placement = "outside", strip.text.y = element_text(angle = 0),
          axis.title.y = element_blank()) }

#For all groups at the same time

pfoa_lc_pr2 <- plot_adj_result2(data = results_lc_pr2, pfas = "conc_pfoa") + 
  theme(strip.placement = "outside")
pfos_lc_pr2 <- plot_adj_result2(data = results_lc_pr2, pfas = "conc_pfos") + 
  theme(axis.text.y = element_blank(), axis.ticks.y = element_blank(), strip.text.y = element_blank(), 
        legend.position = "bottom")
pfna_lc_pr2 <- plot_adj_result2(data = results_lc_pr2, pfas = "conc_pfna") + 
  theme(axis.text.y = element_blank(), axis.ticks.y = element_blank(),  strip.text.y = element_blank())
pfhxs_lc_pr2 <- plot_adj_result2(data = results_lc_pr2, pfas = "conc_pfhxs") +
  theme(axis.text.y = element_blank(), axis.ticks.y = element_blank(),  strip.text.y = element_blank())

# Combine plots using patchwork
combined_plot2 <- pfoa_lc_pr2 | pfos_lc_pr2 | pfna_lc_pr2 | pfhxs_lc_pr2

combined_plot2

#Cyclooxygenase only

pfoa_lc_pr2_cox <- plot_adj_result2(data = results_lc_pr2, pfas = "conc_pfoa", subset = "Cyclooxygenase", width = 0.5) + 
  theme(strip.placement = "outside")
pfos_lc_pr2_cox <- plot_adj_result2(data = results_lc_pr2, pfas = "conc_pfos", subset = "Cyclooxygenase", width = 0.5) + 
  theme(axis.text.y = element_blank(), axis.ticks.y = element_blank(), strip.text.y = element_blank(), 
        legend.position = "bottom")
pfna_lc_pr2_cox <- plot_adj_result2(data = results_lc_pr2, pfas = "conc_pfna", subset = "Cyclooxygenase", width = 0.5) + 
  theme(axis.text.y = element_blank(), axis.ticks.y = element_blank(),  strip.text.y = element_blank())
pfhxs_lc_pr2_cox <- plot_adj_result2(data = results_lc_pr2, pfas = "conc_pfhxs", subset = "Cyclooxygenase", width = 0.5) +
  theme(axis.text.y = element_blank(), axis.ticks.y = element_blank(),  strip.text.y = element_blank())

# Combine plots using patchwork
combined_plot2_cox <- pfoa_lc_pr2_cox | pfos_lc_pr2_cox | pfna_lc_pr2_cox | pfhxs_lc_pr2_cox

combined_plot2_cox

#C450 only

pfoa_lc_pr2_c450 <- plot_adj_result2(data = results_lc_pr2, pfas = "conc_pfoa", subset = "Cytochrome p450", width = 0.5) + 
  theme(strip.placement = "outside")
pfos_lc_pr2_c450 <- plot_adj_result2(data = results_lc_pr2, pfas = "conc_pfos", subset = "Cytochrome p450", width = 0.5) + 
  theme(axis.text.y = element_blank(), axis.ticks.y = element_blank(), strip.text.y = element_blank(), 
        legend.position = "bottom")
pfna_lc_pr2_c450 <- plot_adj_result2(data = results_lc_pr2, pfas = "conc_pfna", subset = "Cytochrome p450", width = 0.5) + 
  theme(axis.text.y = element_blank(), axis.ticks.y = element_blank(),  strip.text.y = element_blank())
pfhxs_lc_pr2_c450 <- plot_adj_result2(data = results_lc_pr2, pfas = "conc_pfhxs", subset = "Cytochrome p450", width = 0.5) +
  theme(axis.text.y = element_blank(), axis.ticks.y = element_blank(),  strip.text.y = element_blank())

# Combine plots using patchwork
combined_plot2_c450 <- pfoa_lc_pr2_c450 | pfos_lc_pr2_c450 | pfna_lc_pr2_c450 | pfhxs_lc_pr2_c450

combined_plot2_c450


#Lipoxygenase only

pfoa_lc_pr2_lox <- plot_adj_result2(data = results_lc_pr2, pfas = "conc_pfoa", subset = "Lipoxygenase", width = 0.5) + 
  theme(strip.placement = "outside")
pfos_lc_pr2_lox <- plot_adj_result2(data = results_lc_pr2, pfas = "conc_pfos", subset = "Lipoxygenase", width = 0.5) + 
  theme(axis.text.y = element_blank(), axis.ticks.y = element_blank(), strip.text.y = element_blank(), 
        legend.position = "bottom")
pfna_lc_pr2_lox <- plot_adj_result2(data = results_lc_pr2, pfas = "conc_pfna", subset = "Lipoxygenase", width = 0.5) + 
  theme(axis.text.y = element_blank(), axis.ticks.y = element_blank(),  strip.text.y = element_blank())
pfhxs_lc_pr2_lox <- plot_adj_result2(data = results_lc_pr2, pfas = "conc_pfhxs", subset = "Lipoxygenase", width = 0.5) +
  theme(axis.text.y = element_blank(), axis.ticks.y = element_blank(),  strip.text.y = element_blank())

# Combine plots using patchwork
combined_plot2_lox <- pfoa_lc_pr2_lox | pfos_lc_pr2_lox | pfna_lc_pr2_lox | pfhxs_lc_pr2_lox

combined_plot2_lox



#Parent compounds only

pfoa_lc_pr2_parent <- plot_adj_result2(data = results_lc_pr2, pfas = "conc_pfoa", subset = "Parent Compound", width = 0.5) + 
  theme(strip.placement = "outside")
pfos_lc_pr2_parent <- plot_adj_result2(data = results_lc_pr2, pfas = "conc_pfos", subset = "Parent Compound", width = 0.5) + 
  theme(axis.text.y = element_blank(), axis.ticks.y = element_blank(), strip.text.y = element_blank(), 
        legend.position = "bottom")
pfna_lc_pr2_parent <- plot_adj_result2(data = results_lc_pr2, pfas = "conc_pfna", subset = "Parent Compound", width = 0.5) + 
  theme(axis.text.y = element_blank(), axis.ticks.y = element_blank(),  strip.text.y = element_blank())
pfhxs_lc_pr2_parent <- plot_adj_result2(data = results_lc_pr2, pfas = "conc_pfhxs", subset = "Parent Compound", width = 0.5) +
  theme(axis.text.y = element_blank(), axis.ticks.y = element_blank(),  strip.text.y = element_blank())

# Combine plots using patchwork
combined_plot2_parent <- pfoa_lc_pr2_parent | pfos_lc_pr2_parent | pfna_lc_pr2_parent | pfhxs_lc_pr2_parent

combined_plot2_parent


```


### Adjusted for weights in LC and rerun the model 

```{r adjust for weights in LC and rerun the model}

pfas_eic_lcwt2 <- pfas_eic_lc %>% 
  mutate(phase1_weight = ifelse(ptb == 1, 1.1, 2.95)) %>% 
  filter(!is.na(ptb))

# Run the function for LC
results_lcwt <- pfas_eic_adjmod(data = pfas_eic_lcwt2, exposures = exposures_lc,  
                              outcomes = eic_all2, covariates = covars_list_lc, weg = "phase1_weight")

results_lcwt2 <- results_lcwt %>% mutate(vars = Outcome) %>%
  inner_join(eic_class, by = "vars") %>% select(-vars)

names(results_lcwt2) <- c("exposure", "outcome", "estimate", "conf.low", "conf.high", "pval", "group")


pfoa_lcwt<- plot_adj_result(data = results_lcwt2, pfas = "conc_pfoa") + theme(strip.placement = "outside") + ggtitle("Associations in LC")
pfos_lcwt<- plot_adj_result(data = results_lcwt2, pfas = "conc_pfos") + theme(axis.text.y = element_blank(), axis.ticks.y = element_blank(), strip.text.y = element_blank())
pfna_lcwt<- plot_adj_result(data = results_lcwt2, pfas = "conc_pfna") + theme(axis.text.y = element_blank(), axis.ticks.y = element_blank(),  strip.text.y = element_blank())
pfhxs_lcwt<- plot_adj_result(data = results_lcwt2, pfas = "conc_pfhxs") + theme(axis.text.y = element_blank(), axis.ticks.y = element_blank(),  strip.text.y = element_blank())
pfua_lcwt<- plot_adj_result(data = results_lcwt2, pfas = "conc_pfua") +  theme(axis.text.y = element_blank(), axis.ticks.y = element_blank(),  strip.text.y = element_blank())
pfdea_lcwt<- plot_adj_result(data = results_lcwt2, pfas = "conc_pfdea") + theme(axis.text.y = element_blank(), axis.ticks.y = element_blank(),  strip.text.y = element_blank())
mpah_lcwt<- plot_adj_result(data = results_lcwt2, pfas = "conc_me.pfosa.acoh") + theme(axis.text.y = element_blank(), axis.ticks.y = element_blank(),  strip.text.y = element_blank())

# combine plots using patchwork
lc_plotwt <-  pfoa_lcwt | pfos_lcwt | pfna_lcwt | pfhxs_lcwt | pfua_lcwt | pfdea_lcwt | mpah_lcwt

lc_plotwt

```

Correlation between beta estimates from weighted vs unweighted models for PR

```{r correlation of beta estimates from weighted vs unweighted LC models}

beta_chk <- results_lc2 %>% inner_join(results_lcwt2, by = c("outcome", "exposure"))

beta_chk %>% ggplot(aes(x = estimate.x, y = estimate.y, col = exposure)) +
  geom_point() +
  facet_wrap(~group.x, scales = "fixed") +
  xlab("Unweighted beta") + ylab("Weighted beta") +
  geom_abline(intercept = 0, slope = 1) +
  coord_fixed(ratio = 1) +
  ggtitle("Comparing betas")



beta_chk %>% ggplot(aes(x = pval.x, y = pval.y, col = exposure)) +
  geom_point() +
  facet_wrap(~group.x, scales = "fixed") +
  xlab("pval-unweighted") + ylab("pval-weighted") +
  geom_abline(intercept = 0, slope = 1) +
  coord_fixed(ratio = 1) +
  ggtitle("Comparing p-values") +
  theme(legend.position = "none") +
  geom_vline(xintercept = 0.05) +
  geom_hline(yintercept = 0.05)

cor(beta_chk$estimate.x, beta_chk$estimate.y, method = "pearson")
  

```



### Select associations that are significant in at least one cohort

```{r now we will select associations that are significant in both LC and PR}

# Filter for significant associations in either cohort
sig_ass_lc_pr20 <- results_lc_pr2  %>%
  filter(pval < 0.1) %>% select(exposure, outcome)

sig_ass_lc_pr2 <- sig_ass_lc_pr20 %>% inner_join(results_lc_pr2, by = c("exposure", "outcome"))

# Print the significant associations
print(sig_ass_lc_pr2)


pfoa_lc_pr2_sig <- plot_adj_result2(data = sig_ass_lc_pr2, pfas = "conc_pfoa") + 
  theme(strip.placement = "outside")
pfos_lc_pr2_sig <- plot_adj_result2(data = sig_ass_lc_pr2, pfas = "conc_pfos") 
pfna_lc_pr2_sig <- plot_adj_result2(data = sig_ass_lc_pr2, pfas = "conc_pfna") 
pfhxs_lc_pr2_sig <- plot_adj_result2(data = sig_ass_lc_pr2, pfas = "conc_pfhxs") 

# Combine plots using patchwork
ggarrange(pfoa_lc_pr2_sig, pfos_lc_pr2_sig, pfna_lc_pr2_sig, pfhxs_lc_pr2_sig,
          nrow = 1, align = "v", common.legend = TRUE)


```


### Run PCA on eicosanoids in each cohort

```{r PCA on eicosanoids in LC}

eic_lc_only <- pfas_eic_lc %>% select(id, all_of(eic_all2)) 


# Exclude the 'id' column for transformation
eic_data <- eic_lc_only[, -1]

# Apply log transformation (adding a small constant to avoid log(0))
log_transformed <- log(eic_data)

# Standardize the data (mean = 0, variance = 1)
scaled_data <- scale(log_transformed)

# Perform PCA
pca_result <- prcomp(scaled_data, center = FALSE, scale. = FALSE)

# View PCA summary
summary(pca_result)

# Scree plot to visualize variance explained
screeplot(pca_result, type = "lines", main = "Scree Plot")

# Biplot for visualization
biplot(pca_result, scale = 0)

# Get PCA loadings and scores
pca_loadings <- pca_result$rotation
pca_scores <- pca_result$x

# Inspect the first few PCA loadings
head(pca_loadings)

# Inspect the first few PCA scores
head(pca_scores)

```

We need 12 PCs to explain the variability, so maybe I will not do this analysis. 

### Correlation between PFAS and oxidative stress or inflammatory biomarkers

Look at the correlation between PFAS and OS/inflammatory biomarkers in PROTECT

```{r correlation between eicosanoids and OS in PR}

os_pr <- read.csv("os_inf_protect.csv") %>% 
  select(id, visit, method_ind, X_8_IsoM, PGF2a, X_8_Iso) %>%
  pivot_wider(names_from = visit, values_from = c(X_8_IsoM, PGF2a, X_8_Iso))
pfas_eic_pr_os <- pfas_eic_pr %>% left_join(os_pr, by = "id")

#Look at the correlation between eicosanoids and oxidative stress markers

os_pr_list <- c("X_8_IsoM_2", "X_8_IsoM_1", "X_8_IsoM_3", "PGF2a_2", "PGF2a_1", "PGF2a_3",
                "X_8_Iso_2",  "X_8_Iso_1",  "X_8_Iso_3")


cor_eic_osinf_func(data = pfas_eic_pr_os, cohort_variable = "PR", osinf = os_pr_list) 
  
```

In general, we see positive correlation with pro-inflammatory eicosanoids and negative correlation with
anti-inflammatory eicosanoids. 

Now let us look at the correlation between eicosanoids and OS/Inf biomarkers in LIFECODES 

```{r correlation between eicosanoids and OS in LC}

osinf_lc <- read.csv("C:/Users/rsiwa/OneDrive/Michigan_course/Lifecodes/pfas_os_inflammation/dat3_outcomes_long2.csv") %>% 
  select(id, visit, il10:crpro, iso_cor, ohdg_cor) %>%
  filter(visit == 3) 

protein_lc <- read.csv("C:/Users/rsiwa/OneDrive/Michigan_course/Lifecodes/pfas_os_inflammation/dat_protein.csv") %>%
  select(id, nitrotyrosine, chlorotyrosine, dityrosine)

osinf_lc_list <- c("il10", "il1b", "il6", "tnf", "crpro", "iso_cor", "ohdg_cor", "nitrotyrosine", "chlorotyrosine", "dityrosine")

pfas_eic_lc_osinf <- pfas_eic_lc %>% left_join(osinf_lc, by = "id") %>% left_join(protein_lc, by = "id")

cor_eic_osinf_func(data = pfas_eic_lc_osinf, cohort_variable = "LC", osinf = osinf_lc_list) 

```

IL-1B is positively associated with some pro-inflammatory eicosanoids, but also with some anti-inflammatory cytokines. Negative with LA, which is pro-inflammatory, but positive with ALA, SHA, and EPA which are anti-inflammatory. Same - mixed correlation for IL6 as well. 

### Check whether year should be included in the model

```{r check whether year should be included in the model}

# Run the function for LC
results_lc_year <- pfas_eic_adjmod_year(data = pfas_eic_lc, exposures = exposures_lc, 
                              outcomes = eic_all2, covariates = covars_list_lc)


# Run the function for PR
results_pr_year <- pfas_eic_adjmod_year(data = pfas_eic_pr, exposures = exposures_short, 
                                        outcomes = eic_all2, covariates = covars_list_pr)

# Combine results

results_prlc_year2 <- rbind.data.frame((results_lc_year %>% mutate(cohort = "LC")),
                                       (results_pr_year %>% mutate(cohort = "PR")))
  
View(results_prlc_year2)  
  
```

The beta for year seems to be significant for some PFAS and eicosanoids, but lets check the effect on PFAS coefficients when the year is removed as  covariate 

```{r checking the effect of removing year from a model in LC}

#Running models without year as a covariate
covars_list_pr_noyr <- c("age", "bmi_prepreg", "as.factor(mat_edu_cat)", "as.factor(ins_cat)", "nullip")

# Run the function for LC
results_pr_noyr <- pfas_eic_adjmod(data = pfas_eic_pr, exposures = exposures_short, 
                                   outcomes = eic_all2, covariates = covars_list_pr_noyr) %>%
  mutate(cohort = "PR")
names(results_pr_noyr) <- c("exposure", "outcome", "estimate", "conf.low", "conf.high", "pval", "cohort")


covars_list_lc_noyr <- c("age", "bmi_prepreg", "as.factor(mat_edu_cat)", "as.factor(ins_cat)", "race4b", "nullip")

# Run the function for LC
results_lc_noyr <- pfas_eic_adjmod(data = pfas_eic_lc, exposures = exposures_lc, 
                                   outcomes = eic_all2, covariates = covars_list_lc_noyr) %>%
  mutate(cohort = "LC")
names(results_lc_noyr) <- c("exposure", "outcome", "estimate", "conf.low", "conf.high", "pval", "cohort")


results_lcpr_noyr <- rbind.data.frame(results_pr_noyr, results_lc_noyr) %>%
  mutate(vars = Outcome) %>%
  inner_join(eic_class, by = "vars") %>% select(-vars)

names(results_lcpr_noyr) <- c("exposure", "outcome", "estimate", "conf.low", "conf.high", "pval", "cohort", "group")

#Compare results in LC only first

results_lc_noyr0 <- results_lc_noyr %>% select(-cohort) %>% mutate(year = "no")
results_lc20 <- results_lc2 %>% select(-group) %>% mutate(year = "yes")
results_lc_comp <- rbind.data.frame(results_lc_noyr0, results_lc20)

ggplot(results_lc_comp, aes(x = estimate, xmin = conf.low, xmax = conf.high, y = outcome, col = year)) +
  geom_pointrange(position = position_dodge(width = 0.5), fatten = 1.5) +
  facet_wrap(~ exposure, nrow = 1) + geom_vline(xintercept = 0)

#Then compare results in PR only

results_pr_noyr0 <- results_pr_noyr %>% select(-cohort) %>% mutate(year = "no")
results_pr20 <- results_pr2  %>% select(-group) %>% mutate(year = "yes")
results_pr_comp <- rbind.data.frame(results_pr_noyr0, results_pr20)

ggplot(results_pr_comp, aes(x = estimate, xmin = conf.low, xmax = conf.high, y = outcome, col = year)) +
  geom_pointrange(position = position_dodge(width = 0.5), fatten = 1.5) +
  facet_wrap(~ exposure, nrow = 1) + geom_vline(xintercept = 0)


```

Except for few cases (PGD2, PGD3), the effect of removing year as a covariate in LC models is minimal and does not really change the coefficients of PFAS. On the other hand, the removal of year seems to have a larger impact on PFAS covariates in PROTECT. 


### Multiple correction of p-values

```{r p-values after multiple correction}

#PR

results_pr2b <- results_pr2 %>% mutate(ad_pval = stats::p.adjust(pval, method = "BH"))
results_lc2b <- results_lc2 %>% mutate(ad_pval = stats::p.adjust(pval, method = "BH"))


```

### Adjust for GA at sample collection for eicosanoids [to compare results with Suthar et al.]

```{r adjusting for GA at sample collection for eicosanoids}

covars_list_lc_ga <- c(covars_list_lc, "ga_t3")
covars_list_pr_ga <- c(covars_list_pr, "ga_t3")


# Run the function for LC
results_lc_ga <- pfas_eic_adjmod(data = pfas_eic_lc, exposures = exposures_lc, 
                              outcomes = eic_all2, covariates = covars_list_lc_ga)

results_lc_ga2 <- results_lc_ga %>% mutate(vars = Outcome) %>%
  inner_join(eic_class, by = "vars") %>% select(-vars)

names(results_lc_ga2) <- c("exposure", "outcome", "estimate", "conf.low", "conf.high", "pval", "group")


pfoa_lc_ga<- plot_adj_result(data = results_lc_ga2, pfas = "conc_pfoa") + theme(strip.placement = "outside") + ggtitle("Associations in LC")
pfos_lc_ga<- plot_adj_result(data = results_lc_ga2, pfas = "conc_pfos") + theme(axis.text.y = element_blank(), axis.ticks.y = element_blank(), strip.text.y = element_blank())
pfna_lc_ga<- plot_adj_result(data = results_lc_ga2, pfas = "conc_pfna") + theme(axis.text.y = element_blank(), axis.ticks.y = element_blank(),  strip.text.y = element_blank())
pfhxs_lc_ga<- plot_adj_result(data = results_lc_ga2, pfas = "conc_pfhxs") + theme(axis.text.y = element_blank(), axis.ticks.y = element_blank(),  strip.text.y = element_blank())
pfua_lc_ga<- plot_adj_result(data = results_lc_ga2, pfas = "conc_pfua") +  theme(axis.text.y = element_blank(), axis.ticks.y = element_blank(),  strip.text.y = element_blank())
pfdea_lc_ga<- plot_adj_result(data = results_lc_ga2, pfas = "conc_pfdea") + theme(axis.text.y = element_blank(), axis.ticks.y = element_blank(),  strip.text.y = element_blank())
mpah_lc_ga<- plot_adj_result(data = results_lc_ga2, pfas = "conc_me.pfosa.acoh") + theme(axis.text.y = element_blank(), axis.ticks.y = element_blank(),  strip.text.y = element_blank())

# combine plots using patchwork
lc_plot_ga <-  pfoa_lc_ga | pfos_lc_ga | pfna_lc_ga | pfhxs_lc_ga | pfua_lc_ga | pfdea_lc_ga | mpah_lc_ga

lc_plot_ga


# Run the function for PR
results_pr_ga <- pfas_eic_adjmod(data = pfas_eic_pr, exposures = exposures_short, 
                              outcomes = eic_all2, covariates = covars_list_pr_ga)

results_pr_ga2 <- results_pr_ga %>% mutate(vars = Outcome) %>%
  inner_join(eic_class, by = "vars") %>% select(-vars)

names(results_pr_ga2) <- c("exposure", "outcome", "estimate", "conf.low", "conf.high", "pval", "group")


pfoa_pr_ga<- plot_adj_result(data = results_pr_ga2, pfas = "conc_pfoa") + theme(strip.placement = "outside") + ggtitle("Associations in PR")
pfos_pr_ga<- plot_adj_result(data = results_pr_ga2, pfas = "conc_pfos") + theme(axis.text.y = element_blank(), axis.ticks.y = element_blank(), strip.text.y = element_blank())
pfna_pr_ga<- plot_adj_result(data = results_pr_ga2, pfas = "conc_pfna") + theme(axis.text.y = element_blank(), axis.ticks.y = element_blank(),  strip.text.y = element_blank())
pfhxs_pr_ga<- plot_adj_result(data = results_pr_ga2, pfas = "conc_pfhxs") + theme(axis.text.y = element_blank(), axis.ticks.y = element_blank(),  strip.text.y = element_blank())

# combine plots using patchwork
pr_plot_ga <- pfoa_pr_ga | pfos_pr_ga | pfna_pr_ga | pfhxs_pr_ga 

pr_plot_ga


results_lc_pr_ga2 <- rbind.data.frame((results_pr_ga2 %>% mutate(cohort = "PR")),
                                   (results_lc_ga2 %>% mutate(cohort = "LC")))


#For all groups at the same time

pfoa_lc_pr_ga2 <- plot_adj_result2(data = results_lc_pr_ga2, pfas = "conc_pfoa") + 
  theme(strip.placement = "outside")
pfos_lc_pr_ga2 <- plot_adj_result2(data = results_lc_pr_ga2, pfas = "conc_pfos") + 
  theme(axis.text.y = element_blank(), axis.ticks.y = element_blank(), strip.text.y = element_blank(), 
        legend.position = "bottom")
pfna_lc_pr_ga2 <- plot_adj_result2(data = results_lc_pr_ga2, pfas = "conc_pfna") + 
  theme(axis.text.y = element_blank(), axis.ticks.y = element_blank(),  strip.text.y = element_blank())
pfhxs_lc_pr_ga2 <- plot_adj_result2(data = results_lc_pr_ga2, pfas = "conc_pfhxs") +
  theme(axis.text.y = element_blank(), axis.ticks.y = element_blank(),  strip.text.y = element_blank())

# Combine plots using patchwork
combined_plot_ga2 <- pfoa_lc_pr_ga2 | pfos_lc_pr_ga2 | pfna_lc_pr_ga2 | pfhxs_lc_pr_ga2

combined_plot_ga2


```








