---
title: "PFAS and Eicosanoids"
output: html_notebook
---

This notebook will include analysis related to PFAS and Eicosanoids in both PROTECT only

### Import data 

Load some libraries

```{r loading libraries}

library(dplyr)
library(tidyr)
library(ggplot2)
library(naniar)
library(corrplot)
library(patchwork)
library(ggpubr)
library(pheatmap)
library(forcats)
library(kableExtra)
library(flextable)


```

Import some data files 

```{r import files}

lod_eic <- read.csv("lod_eic_updated.csv") %>% select(analyte,  analyte_format, LOD)

pfas_eic_combined <- read.csv("pfas_eic_pr_lc_comb.csv", check.names = FALSE)[,-1] 
covars_pr <- read.csv("covars_pr.csv")%>% 
  select(-X) 
eic_class <- read.csv("eic_class2.csv") %>% 
  select(-X)

names(eic_class) <- c("vars", "group")
# eic_class$vars <- tolower(eic_class$vars)

# PFAS concentration columns
exposures_all<- grep("^conc_", names(pfas_eic_combined), value = TRUE)
exposures_short <- c("conc_pfoa", "conc_pfos", "conc_pfhxs", "conc_pfna")


covars_list <- c("age", "bmi_prepreg", "as.factor(mat_edu_cat)", "as.factor(ins_cat)", "race4b", "nullip", "as.factor(cohort)")
covars_list0 <- c("age", "bmi_prepreg", "mat_edu_cat", "ins_cat", "race4b", "nullip", "cohort")

covars_list_pr <- c("age", "bmi_prepreg", "as.factor(mat_edu_cat)", "as.factor(ins_cat)", "nullip", "year_t1")


#Two eicosanoids ltc4.me and lte4 have 0 as minimim values. This will cause an error for these eicosanoids later on

pfas_eic_combined <- pfas_eic_combined %>%
  mutate(`Leukotriene E4` = ifelse(`Leukotriene E4` == 0, NA, `Leukotriene E4`),
         `Leukotriene C4 methyl ester` = ifelse(`Leukotriene C4 methyl ester` == 0, NA, `Leukotriene C4 methyl ester`))

eic_all <- names(pfas_eic_combined[,21:75])

covars_lc_pr <- read.csv("covars_pr_lc_comb.csv") %>% select(-X)


pfas_eic_combined2 <- pfas_eic_combined %>% 
  inner_join(covars_lc_pr, by = c("cohort", "id")) %>% 
  distinct()

# eic_all2 <- setdiff(eic_all, c("dipge2", "dipgj2")) #thi was done before because LC cohort doesnt have these compounds. Not necessary for PR

eic_all2 <- eic_all

#Add third visit ga
tvdate_ga <- read.csv("thirdvisitdate.csv") %>% 
  select(-"X") %>% distinct()

pfas_eic_combined2 <- pfas_eic_combined2 %>% 
  left_join(tvdate_ga, by = c("cohort", "id")) 

pfas_eic_pr <- pfas_eic_combined2 %>% filter(cohort == "PR")  

pfas_eic_pr <- pfas_eic_pr %>% distinct()

chk <- pfas_eic_pr %>% filter(id == 2) 

```

Recode some variables 

```{r recode some variables now}

pfas_eic_pr <- pfas_eic_pr %>%
  mutate(
    age_group = ifelse(age <= median(age, na.rm = TRUE), "Below Median", "Above Median"),
    bmi_group = case_when(
      bmi_prepreg <= 25 ~ "BMI <= 25",
      bmi_prepreg > 25 & bmi_prepreg <= 30 ~ "25 < BMI <= 30",
      bmi_prepreg > 30 ~ "BMI > 30")
  )

#Recode year
pfas_eic_pr <- pfas_eic_pr %>%
  mutate(year_t1b = ifelse(
    year_t1 %in% 2011:2015, 2015, year_t1)#3 years
  )

pfas_eic_pr <- pfas_eic_pr %>% distinct()

```


### Demographic table - Table 1 for manuscript

I will create demographic table here. This will be table 1 for manuscript. 

```{r creating demographic table for manuscript}

table_output_overall <- table1::table1(~ age + 
                                         bmi_prepreg + 
                                         ga_samp +
                                         ga_t3 +
                                         as.factor(mat_edu_cat) +
                                         as.factor(ins_cat) + 
                                         as.factor(nullip) +
                                         as.factor(smpregb) +
                                         as.factor(year_t1b),
                                       data = pfas_eic_pr)
print(table_output_overall)

table_output_overall_flex <- table1::t1flex(table_output_overall)
save_as_docx(table_output_overall_flex, path = "C:/Users/rsiwa/OneDrive/Michigan_course/eicosanoids//table1_overall.docx")


```


### Distributions of PFAS by cohorts 

First we will look at the distributions of PFAS concentrations in the study participants by cohort

```{r distribution of PFAS by cohort}

pfas_summary_results <- lapply(exposures_all, function(col) {
  pfas_col_summary(pfas_eic_pr, col)
}) %>%
  bind_rows() %>%
  select(PFAS, everything()) %>%
  arrange(desc(Detection_Rate)) %>%
  mutate(MedQ1Q3 = paste0(Median, " (", Q1, ", ", Q3, ")"))

# Print results
print(pfas_summary_results)

# Optional: Save to CSV
write.csv(pfas_summary_results, "pfas_conc_summary_PR.csv", row.names = FALSE)

```

We have 4 PFAS with detection rate above 50% in PR: PFNA, PFHxS, PFOA, and PFOS

Focus on these 4 PFAS for now

```{r correlation between PFAS}

corrdata_pfas <- pfas_eic_pr %>% select(all_of(exposures_short)) %>%
  rename("PFNA" = "conc_pfna", "PFHxS" = "conc_pfhxs", "PFOS" = "conc_pfos", "PFOA" = "conc_pfoa")
create_heatplot(corrdata_pfas, c("PFOA", "PFOS", "PFNA", "PFHxS"), "") + ggtitle("") +
  theme(legend.title = element_blank())

```

In PROTECT, the strongest correlation is between PFOA and PFNA. 

### Correlation between different eicosanoids

Now we will look at correlation between different eicosanoids 

First, in the overall dataset

```{r correlation between eicosanoids in the overall dataset}

# Ensure annotation is sorted by group
eic_class <- eic_class %>% arrange(group)
annotation_row_label <- data.frame(group = as.factor(eic_class$group))
rownames(annotation_row_label) <- eic_class$vars

cor_eic_ls <- cor_eic_func(data = pfas_eic_pr, cohort_variable = "All")
eic_cor <- cor_eic_ls[[1]] 
cor_eic_pr <- cor_eic_ls[[2]]

cor_eic_pr_tab <- as.data.frame(cor_eic_pr) 
cor_eic_pr_tab2 <- cor_eic_pr_tab %>% mutate(var1 = rownames(cor_eic_pr_tab))
cor_eic_pr_tab2_long <- cor_eic_pr_tab2 %>% pivot_longer(cols = `Bicyclo Prostaglandin E1`:`Linoleic Acid`,
                                                        names_to = "var2", values_to = "corr") %>%
  filter(var1 != var2) %>% inner_join(eic_class, by = c("var1" = "vars")) %>%
  inner_join(eic_class, by = c("var2" = "vars")) %>%
  mutate(signcorr = sign(corr))

within_path_cor <- cor_eic_pr_tab2_long %>% group_by(group.x, group.y, signcorr) %>% summarise(mean = mean(corr))

```

### Get the most important correlation [eho > abs(0.5)]

```{r choosing the higher correlation}

cor_to_choose_pr <- which(cor_eic_pr >= 0|cor_eic_pr <=0, arr.ind = TRUE)

cor_to_choose_pr2 <- data.frame(
  Variable1 = rownames(cor_eic_pr)[cor_to_choose_pr[, 1]],
  Variable2 = colnames(cor_eic_pr)[cor_to_choose_pr[, 2]],
  Correlation = cor_eic_pr[cor_to_choose_pr]
)

cor_eic_pr3 <- cor_eic_func2(data = pfas_eic_pr, cohort_variable = "PR", threshold = 0.5)

cor_to_choose_pr2b <- cor_to_choose_pr2 %>% filter(Variable1 != Variable2)

```

Check the distributions of eicosanoids in LC and PR

```{r distributions of eicosanoids in LC and PR}

pfas_eic_pr_long <- pfas_eic_pr %>% select(id, cohort, 
                                           `13,14-dihydro-15-keto Prostaglandin F2α`:`13,14-dihydro-15-keto Prostaglandin J2`) %>%
  pivot_longer(cols = `13,14-dihydro-15-keto Prostaglandin F2α`:`13,14-dihydro-15-keto Prostaglandin J2`, 
               names_to = "vars", values_to = "eic_value") %>%
  inner_join(eic_class, by = "vars")

eic_hist_fun <- function(group_name) {
  p <- pfas_eic_pr_long %>%
    filter(group == group_name) %>%
    ggplot(aes(x = log(eic_value), fill = group_name)) +
    geom_density() +
    facet_wrap(~vars, nrow = 3, scales = "free") +
    ggtitle(paste(group_name, "group")) +
    theme(
      legend.position = "none",
      strip.text = element_text(size = 8) # Adjust the text size for facet labels
    ) +
    xlab("log(concentration)")
  return(p)
}


# Now call the function
p1eic <- eic_hist_fun(group_name = "Cyclooxygenase")
p2eic <- eic_hist_fun(group_name = "Cytochrome p450")
p3eic <- eic_hist_fun(group_name = "Lipoxygenase")
p4eic <- eic_hist_fun(group_name = "Parent Compound")

p1eic
p2eic
p3eic
p4eic

```

Check eicosanoids with negative values 

```{r check eiocsanoids with negative values}
summary(pfas_eic_pr)

```

The minimum value for 20HETE is negative. 

Generally, eicosanoid levels are higher in PROTECT as compared to LIFECODES

### Obtain the distributions of eicosanoids in PROTECT

```{r obtaining the distributions of eicosanoids in PROTECT}

eic_summary_results <- lapply(eic_all2, function(col) {
  eic_col_summary(pfas_eic_pr, col)
}) %>%
  bind_rows() %>%
  select(Eicosanoids, everything()) %>%
  mutate(MedQ1Q3 = paste0(Median, " (", Q1, ", ", Q3, ")"), vars = Eicosanoids)

# Add class labels

eic_summary_results2 <- eic_summary_results %>% 
  inner_join(eic_class, by = "vars") %>%
  select(group, Eicosanoids, Median, Q1, Q3, MedQ1Q3)

# Optional: Save to CSV
write.csv(eic_summary_results2, "eic_conc_summary_PR.csv", row.names = FALSE)


```

I want to now look what % of samples were below the LOD for each eicosanoid

This can be done by first converting eicosanoids data to long-form and then using the LOD to determine the det rate

```{r detection rate of eicosanoids}

pfas_eic_pr_long <- pfas_eic_pr %>% select(id, all_of(eic_all2)) %>%
  pivot_longer(cols = `13,14-dihydro-15-keto Prostaglandin F2α`:`13,14-dihydro-15-keto Prostaglandin J2`,
               names_to = "analyte_format", values_to = "conc")
# pfas_eic_pr_long$analyte <- tolower(pfas_eic_pr_long$analyte)

pfas_eic_pr_long2 <- pfas_eic_pr_long %>% inner_join(lod_eic, by = "analyte_format")

pfas_eic_pr_long2_notNA <- pfas_eic_pr_long2 %>% filter(!is.na(LOD)) %>%
  mutate(BLOD = ifelse(conc < LOD, 1, 0))

detrate <- pfas_eic_pr_long2_notNA %>% select(analyte_format, BLOD) %>%
  group_by(analyte_format) %>% summarise(detrate = (1-mean(BLOD, na.rm=TRUE))*100) %>% arrange(desc(detrate))

chk <- pfas_eic_pr %>% select(id, `12(S)-HETE`, `(±)8,9-DHET`, `Prostaglandin E2`, `Leukotriene E4`, 
                              `Leukotriene C4 methyl ester`,`15(S)-HETE`)

```

These compounds have det rate below 70%: `12(S)-HETE`, `(±)8,9-DHET`, `Prostaglandin E2`, `Leukotriene E4`, 
                              `Leukotriene C4 methyl ester`,`15(S)-HETE`

### Correlation between PFAS and eicosanoids 

```{r correlation between PFAS and eicosanoids}

annotation_row_label2 <- data.frame(group = as.factor(eic_class$group))
rownames(annotation_row_label2) <- eic_class$vars

corrdata_pfas_eic <- cor_eic_pfas_func(data = pfas_eic_pr, cohort_variable = "PR",
                   pfas_vec = exposures_short)
corrdata_pfas_eic[[2]]

```

### Adjusted models for PFAS and Eicosanoids 

Now we will model the relationship between eicosanoids and PFAS taking into consideration the covariates 

```{r adjusted associations between PFAS and Eicosanoids - PR only}

#For reference
#exposures_short: "conc_pfoa"  "conc_pfos"  "conc_pfhxs" "conc_pfna" 
#covars_list_pr: "age" "bmi_prepreg" "as.factor(mat_edu_cat)" "as.factor(ins_cat)" "nullip" "year_t1"

pfas_eic_pr2 <- pfas_eic_pr
names(pfas_eic_pr2) <- make.names(names(pfas_eic_pr)) #to use when i get error with the current names
eic_all2b <- make.names(eic_all2) #to use when i get error with the current names
eic_class2 <- eic_class %>% rename(vars_format = vars)
eic_class2$vars <- make.names(eic_class2$vars_format)

# Run the function for PR
results_pr <- pfas_eic_adjmod(data = pfas_eic_pr2, exposures = exposures_short, 
                                outcomes = eic_all2b, covariates = covars_list_pr)

results_pr2 <- results_pr %>% mutate(vars = Outcome) %>%
  inner_join(eic_class2, by = "vars") %>% select(-vars)

names(results_pr2) <- c("exposure", "outcome", "estimate", "conf.low", "conf.high", "pval", "outcome_format", "group")

#convert results in terms of IQR increase in PFAS

pfas_iqr <- pfas_eic_pr %>%
  select(all_of(exposures_short)) %>%  # Select PFAS columns based on exposures_short
  summarise(across(everything(), ~ IQR(log(.x), na.rm = TRUE)))  # Calculate IQR for each column

# Transpose results for better readability
pfas_iqr <- data.frame(
  exposure = exposures_short,
  IQR = as.numeric(t(pfas_iqr))
)


results_pr3 <- results_pr2 %>%
  rename(estimate_old = estimate, conf.low_old = conf.low, conf.high_old = conf.high) %>%
  inner_join(pfas_iqr, by = "exposure") %>%
  mutate(estimate = (exp(estimate_old*IQR)-1)*100, 
         conf.low = (exp(conf.low_old*IQR)-1)*100, 
         conf.high = (exp(conf.high_old*IQR)-1)*100)


#plot

pfoa_pr<- plot_adj_result(data = results_pr3, pfas = "conc_pfoa") + theme(strip.placement = "outside")
pfos_pr<- plot_adj_result(data = results_pr3, pfas = "conc_pfos") + theme(axis.text.y = element_blank(), axis.ticks.y = element_blank(), strip.text.y = element_blank())
pfna_pr<- plot_adj_result(data = results_pr3, pfas = "conc_pfna") + theme(axis.text.y = element_blank(), axis.ticks.y = element_blank(),  strip.text.y = element_blank())
pfhxs_pr<- plot_adj_result(data = results_pr3, pfas = "conc_pfhxs") + theme(axis.text.y = element_blank(), axis.ticks.y = element_blank(),  strip.text.y = element_blank())

# combine plots using patchwork
pr_plot <- pfoa_pr | pfos_pr | pfna_pr | pfhxs_pr 

pr_plot 


```

### Running model with the same covariates as Suthar et al.

Suthar et al. adjusted for maternal age, education, prepreg BMI, parity, and GA 

```{r run models for eicosanoids in PR with the same covariates as what Suthar et al. did}

covars_list_prb <- c("age", "bmi_prepreg", "as.factor(mat_edu_cat)", "nullip", "ga_t3")

# Run the function for PR
results_prb <- pfas_eic_adjmod(data = pfas_eic_pr2, exposures = exposures_short, 
                                outcomes = eic_all2b, covariates = covars_list_prb)

results_pr2b <- results_prb %>% mutate(vars = Outcome) %>%
  inner_join(eic_class2, by = "vars") %>% select(-vars)

names(results_pr2b) <- c("exposure", "outcome", "estimate", "conf.low", "conf.high", "pval", "outcome_format", "group")


results_pr3b <- results_pr2b %>%
  rename(estimate_old = estimate, conf.low_old = conf.low, conf.high_old = conf.high) %>%
  inner_join(pfas_iqr, by = "exposure") %>%
  mutate(estimate = (exp(estimate_old*IQR)-1)*100, 
         conf.low = (exp(conf.low_old*IQR)-1)*100, 
         conf.high = (exp(conf.high_old*IQR)-1)*100)


#plot

pfoa_prb <- plot_adj_result(data = results_pr3b, pfas = "conc_pfoa") + theme(strip.placement = "outside")
pfos_prb <- plot_adj_result(data = results_pr3b, pfas = "conc_pfos") + theme(axis.text.y = element_blank(), axis.ticks.y = element_blank(), strip.text.y = element_blank())
pfna_prb <- plot_adj_result(data = results_pr3b, pfas = "conc_pfna") + theme(axis.text.y = element_blank(), axis.ticks.y = element_blank(),  strip.text.y = element_blank())
pfhxs_prb <- plot_adj_result(data = results_pr3b, pfas = "conc_pfhxs") + theme(axis.text.y = element_blank(), axis.ticks.y = element_blank(),  strip.text.y = element_blank())

# combine plots using patchwork
pr_plotb <- pfoa_prb | pfos_prb | pfna_prb | pfhxs_prb 

pr_plotb


```

Add year to the previous model and rerun

```{r adding year to the suthar et al model}

covars_list_prc <- c("age", "bmi_prepreg", "as.factor(mat_edu_cat)", "nullip", "ga_t3", "as.numeric(year_t1)")

# Run the function for PR
results_prc <- pfas_eic_adjmod(data = pfas_eic_pr2, exposures = exposures_short, 
                               outcomes = eic_all2b, covariates = covars_list_prc)

results_pr2c <- results_prc %>% mutate(vars = Outcome) %>%
  inner_join(eic_class2, by = "vars") %>% select(-vars)

names(results_pr2c) <- c("exposure", "outcome", "estimate", "conf.low", "conf.high", "pval", "outcome_format", "group")


results_pr3c <- results_pr2c %>%
  rename(estimate_old = estimate, conf.low_old = conf.low, conf.high_old = conf.high) %>%
  inner_join(pfas_iqr, by = "exposure") %>%
  mutate(estimate = (exp(estimate_old*IQR)-1)*100, 
         conf.low = (exp(conf.low_old*IQR)-1)*100, 
         conf.high = (exp(conf.high_old*IQR)-1)*100)


#plot

pfoa_prc <- plot_adj_result(data = results_pr3c, pfas = "conc_pfoa") + theme(strip.placement = "outside")
pfos_prc <- plot_adj_result(data = results_pr3c, pfas = "conc_pfos") + theme(axis.text.y = element_blank(), axis.ticks.y = element_blank(), strip.text.y = element_blank())
pfna_prc <- plot_adj_result(data = results_pr3c, pfas = "conc_pfna") + theme(axis.text.y = element_blank(), axis.ticks.y = element_blank(),  strip.text.y = element_blank())
pfhxs_prc <- plot_adj_result(data = results_pr3c, pfas = "conc_pfhxs") + theme(axis.text.y = element_blank(), axis.ticks.y = element_blank(),  strip.text.y = element_blank())

# combine plots using patchwork
pr_plotc <- pfoa_prc | pfos_prc | pfna_prc | pfhxs_prc 

pr_plotc

```

Verifying the results from the function

```{r verifying the results from the function}

covars_list_pr
covars_list_prb
covars_list_prc

med_gat3 <- median(pfas_eic_pr2$ga_t3, na.rm = TRUE)

pfas_eic_pr2 <- pfas_eic_pr2 %>% mutate(ga_t3_imp = ifelse(is.na(ga_t3), med_gat3, ga_t3))

chk_eic <- lm(log(Linoleic.Acid) ~ log(conc_pfos) + age + bmi_prepreg + as.factor(mat_edu_cat) + 
                as.factor(ins_cat) + nullip + year_t1, data = pfas_eic_pr2)
broom::tidy(chk_eic)
car::vif(chk_eic)


chk_eicb0 <- lm(log(Linoleic.Acid) ~ log(conc_pfos) + age + bmi_prepreg + as.factor(mat_edu_cat) + 
                  nullip + year_t1, data = pfas_eic_pr2)
broom::tidy(chk_eicb0)
car::vif(chk_eicb0)


chk_eicb <- lm(log(Linoleic.Acid) ~ log(conc_pfos) + age + bmi_prepreg + as.factor(mat_edu_cat) + 
                nullip + ga_t3_imp, data = pfas_eic_pr2)
broom::tidy(chk_eicb)
car::vif(chk_eicb)


chk_eicc <- lm(log(Linoleic.Acid) ~ log(conc_pfos) + age + bmi_prepreg + as.factor(mat_edu_cat) + 
                nullip + ga_t3 + year_t1, data = pfas_eic_pr2)
broom::tidy(chk_eicc)
car::vif(chk_eicc)

#residual plots

plot(chk_eic)
plot(chk_eicb)
plot(chk_eicc)


chk_eic2 <- lm(log(X.α..Linolenic.Acid) ~ log(conc_pfos) + age + bmi_prepreg + as.factor(mat_edu_cat) + 
                as.factor(ins_cat) + nullip + year_t1, data = pfas_eic_pr2)
broom::tidy(chk_eic2)


chk_eicb2 <- lm(log(X.α..Linolenic.Acid) ~ log(conc_pfos) + age + bmi_prepreg + as.factor(mat_edu_cat) + 
                nullip + ga_t3, data = pfas_eic_pr2)
broom::tidy(chk_eicb2)


chk_eicc2 <- lm(log(X.α..Linolenic.Acid) ~ log(conc_pfos) + age + bmi_prepreg + as.factor(mat_edu_cat) + 
                nullip + ga_t3 + year_t1, data = pfas_eic_pr2)
broom::tidy(chk_eicc2)


plot(chk_eic2)
plot(chk_eicb2)
plot(chk_eicc2)
```

The results look fine, per manual verification. Also adjusting for year provides a better model fit. I think we should adjust for the year because of how importantly it affects the results. 

Check the trend for some eicosanoids with year 

### Outcome trends with year

```{r look at the trend of outcomes with year}


# Reshape the data for plotting
heatmap_data <- pfas_eic_pr2 %>%
  select(all_of(eic_all2b), year_t1) %>%  # Select eicosanoids and year
  group_by(year_t1) %>%                    # Group by year
  summarise(across(everything(), function (x) median(log(x), na.rm = TRUE))) %>%  # Calculate mean for each eicosanoid by year
  pivot_longer(-year_t1, names_to = "Eicosanoid", values_to = "Median")  # Reshape to long format

# Plot the heatmap
heatmap_plot <- ggplot(heatmap_data, aes(x = factor(year_t1), y = Eicosanoid, fill = Median)) +
  geom_tile(color = "white") +  
  scale_fill_gradient() +  # Gradient fill
  labs(x = "Year", y = "Eicosanoid", fill = "Median", title = "Eicosanoids Levels by Year") +
  theme_minimal() +  # Minimal theme
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),  # Rotate x-axis labels for readability
    axis.text.y = element_text(size = 8),              # Adjust y-axis text size
    strip.text = element_text(size = 10)               # Adjust facet strip text size if needed
  )

# Print the plot
print(heatmap_plot)


#Also plot a lineplot
lineplot <- ggplot(heatmap_data, aes(x = as.factor(year_t1), y = Median, color = Eicosanoid, group = Eicosanoid)) +
  geom_line() +  # Add lines
  geom_point(size = 2) +  # Add points for each data value
  labs(
    x = "Year",
    y = "Mean Level",
    color = "Eicosanoid",
    title = "Trends in Eicosanoids Levels by Year"
  ) +
  theme_minimal() +  # Minimal theme
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),  # Rotate x-axis labels for readability
    legend.position = "bottom",                          # Position legend to the right
    legend.text = element_text(size = 8),               # Adjust legend text size
    legend.title = element_text(size = 10),             # Adjust legend title size
    axis.text.y = element_text(size = 8)                # Adjust y-axis text size
  )

print(lineplot)

```

We do not see a consistent trend in eicosanoids


### Analysis with categorized exposures for PFOA and PFHxS

```{r categorized PFOA and PFHxS exposures}

pfas_eic_pr2 <- pfas_eic_pr2 %>% 
  mutate(alod_pfoa = 1 - blod_pfoa, alod_pfhxs = 1 - blod_pfhxs, alod_pfua = 1-blod_pfua)

exposures_short_cat <- c("alod_pfoa", "alod_pfhxs", "alod_pfua")


# Run the function for PR
results_pr_cat <- pfas_eic_adjmod(data = pfas_eic_pr2, exposures = exposures_short_cat, 
                                outcomes = eic_all2b, covariates = covars_list_pr)

results_pr_cat2 <- results_pr_cat %>% mutate(vars = Outcome) %>%
  inner_join(eic_class2, by = "vars") %>% select(-vars)

names(results_pr_cat2) <- c("exposure", "outcome", "estimate", "conf.low", "conf.high", "pval", "outcome_format", "group")

#convert results in terms of IQR increase in PFAS

results_pr_cat3 <- results_pr_cat2 %>%
  rename(estimate_old = estimate, conf.low_old = conf.low, conf.high_old = conf.high) %>%
  left_join(pfas_iqr, by = "exposure") %>%
  mutate(estimate = ifelse(substr(exposure,1,4) == "alod", 
                           (exp(estimate_old)-1)*100, (exp(estimate_old*IQR)-1)*100), 
         conf.low = ifelse(substr(exposure,1,4) == "alod",
                           (exp(conf.low_old)-1)*100, (exp(conf.low_old*IQR)-1)*100),
         conf.high = ifelse(substr(exposure,1,4) == "alod",
                            (exp(conf.high_old)-1)*100, (exp(conf.high_old*IQR)-1)*100))


#plot

pfoa_pr_cat <- plot_adj_result(data = results_pr_cat3, pfas = "alod_pfoa") + theme(strip.placement = "outside")
pfhxs_pr_cat <- plot_adj_result(data = results_pr_cat3, pfas = "alod_pfhxs") + theme(axis.text.y = element_blank(), axis.ticks.y = element_blank(), strip.text.y = element_blank())
pfua_pr_cat <- plot_adj_result(data = results_pr_cat3, pfas = "alod_pfua") + theme(axis.text.y = element_blank(), axis.ticks.y = element_blank(),  strip.text.y = element_blank())

# combine plots using patchwork
pr_plot_cat <- pfoa_pr_cat | pfhxs_pr_cat | pfua_pr_cat

pr_plot_cat
  
```


### Run models without education as a covariate

```{r running models without education as a covariate}
#Visualize missingness of all covariates

visdat::vis_miss(pfas_eic_pr2 %>% select(id, age, bmi_prepreg, mat_edu_cat, ins_cat, nullip,
                                         year_t1, ga_t3))


covars_list_pr_noed <- setdiff(covars_list_pr, "as.factor(mat_edu_cat)")


# Run the function for PR
results_pr_noed <- pfas_eic_adjmod(data = pfas_eic_pr2, exposures = exposures_short, 
                                outcomes = eic_all2b, covariates = covars_list_pr_noed)

results_pr2_noed <- results_pr_noed %>% mutate(vars = Outcome) %>%
  inner_join(eic_class2, by = "vars") %>% select(-vars)

names(results_pr2_noed) <- c("exposure", "outcome", "estimate", "conf.low", "conf.high", "pval", "outcome_format", "group")

#convert results in terms of IQR increase in PFAS

results_pr3_noed <- results_pr2_noed %>%
  rename(estimate_old = estimate, conf.low_old = conf.low, conf.high_old = conf.high) %>%
  inner_join(pfas_iqr, by = "exposure") %>%
  mutate(estimate = (exp(estimate_old*IQR)-1)*100, 
         conf.low = (exp(conf.low_old*IQR)-1)*100, 
         conf.high = (exp(conf.high_old*IQR)-1)*100)
results_pr3_noed <- results_pr3_noed %>% mutate(Group2 = "No edu")
results_pr3_chk <- results_pr3 %>% mutate(Group2 = "All primary covariates")

results_pr3_comp <- rbind.data.frame(results_pr3_chk, results_pr3_noed)

#Now plot the comparisons

results_pr3_comp %>%
    ggplot(aes(x = estimate, xmin = conf.low, xmax = conf.high, y = outcome_format, col = Group2)) +
    geom_pointrange(position = position_dodge(width = 0.8)) +
    theme(legend.position = "bottom") +
    geom_vline(xintercept = 0) +
    xlab("") +
    facet_grid(. ~ exposure, scales = "free_x") +  # Facet by group, switch facet labels to y-axis
    theme(strip.placement = "outside", strip.text.y = element_text(angle = 0),
          axis.title.y = element_blank()) 


```

Maybe I should do this with all covariates. I will run without insurance now

### Run models without insurance as a covariate

```{r running models without insurance as a covariate}

covars_list_pr_noins <- setdiff(covars_list_pr, "as.factor(ins_cat)")


# Run the function for PR
results_pr_noins <- pfas_eic_adjmod(data = pfas_eic_pr2, exposures = exposures_short, 
                                   outcomes = eic_all2b, covariates = covars_list_pr_noins)

results_pr2_noins <- results_pr_noins %>% mutate(vars = Outcome) %>%
  inner_join(eic_class2, by = "vars") %>% select(-vars)

names(results_pr2_noins) <- c("exposure", "outcome", "estimate", "conf.low", "conf.high", "pval", "outcome_format", "group")

#convert results in terms of IQR increase in PFAS

results_pr3_noins <- results_pr2_noins %>%
  rename(estimate_old = estimate, conf.low_old = conf.low, conf.high_old = conf.high) %>%
  inner_join(pfas_iqr, by = "exposure") %>%
  mutate(estimate = (exp(estimate_old*IQR)-1)*100, 
         conf.low = (exp(conf.low_old*IQR)-1)*100, 
         conf.high = (exp(conf.high_old*IQR)-1)*100)
results_pr3_noins <- results_pr3_noins %>% mutate(Group2 = "No ins")

results_pr3_comp2 <- rbind.data.frame(results_pr3_chk, results_pr3_noins)

#Now plot the comparisons

results_pr3_comp2 %>%
  ggplot(aes(x = estimate, xmin = conf.low, xmax = conf.high, y = outcome_format, col = Group2)) +
  geom_pointrange(position = position_dodge(width = 0.8)) +
  theme(legend.position = "bottom") +
  geom_vline(xintercept = 0) +
  xlab("") +
  facet_grid(. ~ exposure, scales = "free_x") +  # Facet by group, switch facet labels to y-axis
  theme(strip.placement = "outside", strip.text.y = element_text(angle = 0),
        axis.title.y = element_blank()) 


```

### Run models without age as a covariate

```{r running models without age as a covariate}

covars_list_pr_noage <- setdiff(covars_list_pr, "age")


# Run the function for PR
results_pr_noage <- pfas_eic_adjmod(data = pfas_eic_pr2, exposures = exposures_short, 
                                   outcomes = eic_all2b, covariates = covars_list_pr_noage)

results_pr2_noage <- results_pr_noage %>% mutate(vars = Outcome) %>%
  inner_join(eic_class2, by = "vars") %>% select(-vars)

names(results_pr2_noage) <- c("exposure", "outcome", "estimate", "conf.low", "conf.high", "pval", "outcome_format", "group")

#convert results in terms of IQR increase in PFAS

results_pr3_noage <- results_pr2_noage %>%
  rename(estimate_old = estimate, conf.low_old = conf.low, conf.high_old = conf.high) %>%
  inner_join(pfas_iqr, by = "exposure") %>%
  mutate(estimate = (exp(estimate_old*IQR)-1)*100, 
         conf.low = (exp(conf.low_old*IQR)-1)*100, 
         conf.high = (exp(conf.high_old*IQR)-1)*100)
results_pr3_noage <- results_pr3_noage %>% mutate(Group2 = "No age")

results_pr3_comp3 <- rbind.data.frame(results_pr3_chk, results_pr3_noage)

#Now plot the comparisons

results_pr3_comp3 %>%
  ggplot(aes(x = estimate, xmin = conf.low, xmax = conf.high, y = outcome_format, col = Group2)) +
  geom_pointrange(position = position_dodge(width = 0.8)) +
  theme(legend.position = "bottom") +
  geom_vline(xintercept = 0) +
  xlab("") +
  facet_grid(. ~ exposure, scales = "free_x") +  # Facet by group, switch facet labels to y-axis
  theme(strip.placement = "outside", strip.text.y = element_text(angle = 0),
        axis.title.y = element_blank()) 

```


### Run models without nullip as a covariate

```{r running models without nullip as a covariate}

covars_list_pr_nonullip <- setdiff(covars_list_pr, "nullip")


# Run the function for PR
results_pr_nonullip <- pfas_eic_adjmod(data = pfas_eic_pr2, exposures = exposures_short, 
                                   outcomes = eic_all2b, covariates = covars_list_pr_nonullip)

results_pr2_nonullip <- results_pr_nonullip %>% mutate(vars = Outcome) %>%
  inner_join(eic_class2, by = "vars") %>% select(-vars)

names(results_pr2_nonullip) <- c("exposure", "outcome", "estimate", "conf.low", "conf.high", "pval", "outcome_format", "group")

#convert results in terms of IQR increase in PFAS

results_pr3_nonullip <- results_pr2_nonullip %>%
  rename(estimate_old = estimate, conf.low_old = conf.low, conf.high_old = conf.high) %>%
  inner_join(pfas_iqr, by = "exposure") %>%
  mutate(estimate = (exp(estimate_old*IQR)-1)*100, 
         conf.low = (exp(conf.low_old*IQR)-1)*100, 
         conf.high = (exp(conf.high_old*IQR)-1)*100)
results_pr3_nonullip <- results_pr3_nonullip %>% mutate(Group2 = "No nullip")

results_pr3_comp4 <- rbind.data.frame(results_pr3_chk, results_pr3_nonullip)

#Now plot the comparisons

results_pr3_comp4 %>%
  ggplot(aes(x = estimate, xmin = conf.low, xmax = conf.high, y = outcome_format, col = Group2)) +
  geom_pointrange(position = position_dodge(width = 0.8)) +
  theme(legend.position = "bottom") +
  geom_vline(xintercept = 0) +
  xlab("") +
  facet_grid(. ~ exposure, scales = "free_x") +  # Facet by group, switch facet labels to y-axis
  theme(strip.placement = "outside", strip.text.y = element_text(angle = 0),
        axis.title.y = element_blank()) 

```

### Run models without year as a covariate

```{r running models without year as a covariate}

covars_list_pr_noyear <- setdiff(covars_list_pr, "year_t1")


# Run the function for PR
results_pr_noyear <- pfas_eic_adjmod(data = pfas_eic_pr2, exposures = exposures_short, 
                                   outcomes = eic_all2b, covariates = covars_list_pr_noyear)

results_pr2_noyear <- results_pr_noyear %>% mutate(vars = Outcome) %>%
  inner_join(eic_class2, by = "vars") %>% select(-vars)

names(results_pr2_noyear) <- c("exposure", "outcome", "estimate", "conf.low", "conf.high", "pval", "outcome_format", "group")

#convert results in terms of IQR increase in PFAS

results_pr3_noyear <- results_pr2_noyear %>%
  rename(estimate_old = estimate, conf.low_old = conf.low, conf.high_old = conf.high) %>%
  inner_join(pfas_iqr, by = "exposure") %>%
  mutate(estimate = (exp(estimate_old*IQR)-1)*100, 
         conf.low = (exp(conf.low_old*IQR)-1)*100, 
         conf.high = (exp(conf.high_old*IQR)-1)*100)
results_pr3_noyear <- results_pr3_noyear %>% mutate(Group2 = "No year")

results_pr3_comp5 <- rbind.data.frame(results_pr3_chk, results_pr3_noyear)

#Now plot the comparisons

results_pr3_comp5 %>%
  ggplot(aes(x = estimate, xmin = conf.low, xmax = conf.high, y = outcome_format, col = Group2)) +
  geom_pointrange(position = position_dodge(width = 0.8)) +
  theme(legend.position = "bottom") +
  geom_vline(xintercept = 0) +
  xlab("") +
  facet_grid(. ~ exposure, scales = "free_x") +  # Facet by group, switch facet labels to y-axis
  theme(strip.placement = "outside", strip.text.y = element_text(angle = 0),
        axis.title.y = element_blank()) 

```


### Run models with ga_t3 as a covariate

```{r running models with ga_t3 as a covariate}

covars_list_pr_gat3_added <- c(covars_list_pr, "ga_t3_imp")


# Run the function for PR
results_pr_gat3_added <- pfas_eic_adjmod(data = pfas_eic_pr2, exposures = exposures_short, 
                                   outcomes = eic_all2b, covariates = covars_list_pr_gat3_added)

results_pr2_gat3_added <- results_pr_gat3_added %>% mutate(vars = Outcome) %>%
  inner_join(eic_class2, by = "vars") %>% select(-vars)

names(results_pr2_gat3_added) <- c("exposure", "outcome", "estimate", "conf.low", "conf.high", "pval", "outcome_format", "group")

#convert results in terms of IQR increase in PFAS

results_pr3_gat3_added <- results_pr2_gat3_added %>%
  rename(estimate_old = estimate, conf.low_old = conf.low, conf.high_old = conf.high) %>%
  inner_join(pfas_iqr, by = "exposure") %>%
  mutate(estimate = (exp(estimate_old*IQR)-1)*100, 
         conf.low = (exp(conf.low_old*IQR)-1)*100, 
         conf.high = (exp(conf.high_old*IQR)-1)*100)
results_pr3_gat3_added <- results_pr3_gat3_added %>% mutate(Group2 = "Yes GA_t3")

results_pr3_comp6 <- rbind.data.frame(results_pr3_chk, results_pr3_gat3_added)

#Now plot the comparisons

results_pr3_comp6 %>%
  ggplot(aes(x = estimate, xmin = conf.low, xmax = conf.high, y = outcome_format, col = Group2)) +
  geom_pointrange(position = position_dodge(width = 0.8)) +
  theme(legend.position = "bottom") +
  geom_vline(xintercept = 0) +
  xlab("") +
  facet_grid(group ~ exposure, scales = "free") +  # Facet by group, switch facet labels to y-axis
  theme(strip.placement = "outside", strip.text.y = element_text(angle = 0),
        axis.title.y = element_blank()) 
```

### Now plot results from all models together

Now we will want to plot results from all models 

```{r plot results from all models together}

results_pr3_comp7 <- rbind.data.frame(results_pr3_chk, results_pr3_noed, results_pr3_noins, 
                                      results_pr3_noage, results_pr3_nonullip, results_pr3_noyear,
                                      results_pr3_gat3_added)


plot_allcovarscomp <- function(data, groupvar){
data %>% filter(group == groupvar) %>%
  ggplot(aes(x = estimate, xmin = conf.low, xmax = conf.high, y = outcome_format, col = Group2)) +
  geom_pointrange(position = position_dodge(width = 0.8)) +
  theme(legend.position = "bottom") +
  geom_vline(xintercept = 0) +
  xlab(paste0(groupvar, "group")) +
  facet_grid(.~ exposure, scales = "free") +  # Facet by group, switch facet labels to y-axis
  theme(strip.placement = "outside", strip.text.y = element_text(angle = 0),
        axis.title.y = element_blank()) 
}

plot_allcovarscomp(data = results_pr3_comp7, groupvar = "Cyclooxygenase")
plot_allcovarscomp(data = results_pr3_comp7, groupvar = "Cytochrome p450")
plot_allcovarscomp(data = results_pr3_comp7, groupvar = "Lipoxygenase")
plot_allcovarscomp(data = results_pr3_comp7, groupvar = "Parent Compound")



```
No covariates really move the estimate other than year of enrollment.


### Run PCA on eicosanoids in each cohort

```{r PCA on eicosanoids in PR}

eic_pr_only <- pfas_eic_pr2 %>% select(id, all_of(eic_all2b)) 

#Impute missing data using median

med_lte4 <- median(eic_pr_only$Leukotriene.E4, na.rm = TRUE)
med_ltc4.me <- median(eic_pr_only$Leukotriene.C4.methyl.ester, na.rm = TRUE)
med_hete20 <- median(eic_pr_only$X20.S..HETE, na.rm = TRUE)

eic_pr_only <- eic_pr_only %>%
  mutate(X20.S..HETE = ifelse(X20.S..HETE < 0, NA, X20.S..HETE)) %>%
  mutate(Leukotriene.E4 = ifelse(is.na(Leukotriene.E4), med_lte4, Leukotriene.E4),
         Leukotriene.C4.methyl.ester = ifelse(is.na(Leukotriene.C4.methyl.ester), med_ltc4.me,
                                              Leukotriene.C4.methyl.ester),
         X20.S..HETE = ifelse(is.na(X20.S..HETE), med_hete20, X20.S..HETE))


# Exclude the 'id' column for transformation
eic_data <- eic_pr_only[, -1]

# Apply log transformation (adding a small constant to avoid log(0))
log_transformed <- log(eic_data)

# Standardize the data (mean = 0, variance = 1)
scaled_data <- scale(log_transformed)

# Perform PCA
pca_result <- prcomp(scaled_data, center = FALSE, scale. = FALSE)

# View PCA summary
summary(pca_result)

# Scree plot to visualize variance explained
screeplot(pca_result, type = "lines", main = "Scree Plot")

# Biplot for visualization
biplot(pca_result, scale = 0)

# Get PCA loadings and scores
pca_loadings <- pca_result$rotation
pca_scores <- pca_result$x

# Inspect the first few PCA loadings
head(pca_loadings)

# Inspect the first few PCA scores
head(pca_scores)

```

We need 11 PCs to explain at least 70% variability, so maybe I will not do this analysis. 

### Correlation between PFAS and oxidative stress or inflammatory biomarkers

Look at the correlation between PFAS and OS/inflammatory biomarkers in PROTECT

```{r correlation between eicosanoids and OS in PR}

os_pr <- read.csv("os_inf_protect.csv") %>% 
  select(id, visit, method_ind, X_8_IsoM, PGF2a, X_8_Iso) %>%
  pivot_wider(names_from = visit, values_from = c(X_8_IsoM, PGF2a, X_8_Iso))

pfas_eic_pr_os <- pfas_eic_pr %>% left_join(os_pr, by = "id") %>% distinct()
#563 because for 22 participants, we have OS data from both method1 and method 2 as shown below:


os_pr_id <- pfas_eic_pr_os %>% group_by(id) %>% tally() %>% filter(n == 2)
os_pr_id2 <- os_pr_id %>% inner_join(pfas_eic_pr_os, by = "id") %>%
  select(id, method_ind, all_of(os_pr_list))


#Look at the correlation between eicosanoids and oxidative stress markers

os_pr_list <- c("X_8_IsoM_2", "X_8_IsoM_1", "X_8_IsoM_3", "PGF2a_2", "PGF2a_1", "PGF2a_3",
                "X_8_Iso_2",  "X_8_Iso_1",  "X_8_Iso_3")


cor_eic_osinf_func(data = pfas_eic_pr_os, cohort_variable = "PR", osinf = os_pr_list) 


```

In general, we see positive correlation with pro-inflammatory eicosanoids and negative correlation with
anti-inflammatory eicosanoids. 

### Select only the significant results from the first model

```{r selecting only the significant results}

sig_pfas_eic <-  results_pr3 %>% 
  filter(pval <= 0.05) %>% 
  mutate(sign = ifelse(estimate_old < 0, "Neg", "Pos")) %>%
  select(exposure, outcome, sign) %>%
  pivot_wider(names_from = exposure, values_from = sign) %>% 
  mutate(across(conc_pfoa:conc_pfna, ~ ifelse(is.na(.x), "", .x))) %>% 
  mutate(sign2 = paste0(conc_pfoa, conc_pfhxs, conc_pfos, conc_pfna)) %>%
  mutate(sign3 = substr(sign2, 1,3)) 

sig_pfas_eic_pr <- sig_pfas_eic %>% select(outcome, sign3) %>%
  rename("metab1" = "outcome", "effect sign" = "sign3")
write.csv(sig_pfas_eic_pr, "pfas_eic_pr_sigeffects.csv")

```



### Multiple correction of p-values

I will apply multiple correction for each PFAS 

```{r p-values after multiple correction}

#PR

results_pr3a<- results_pr3 %>% filter(exposure == "conc_pfos") %>%
  mutate(ad_pval = stats::p.adjust(pval, method = "BH"))


results_pr3b <- results_pr3 %>% filter(exposure == "conc_pfna") %>%
  mutate(ad_pval = stats::p.adjust(pval, method = "BH"))


results_pr3c<- results_pr3 %>% filter(exposure == "conc_pfhxs") %>%
  mutate(ad_pval = stats::p.adjust(pval, method = "BH"))


results_pr3d <- results_pr3 %>% filter(exposure == "conc_pfna") %>%
  mutate(ad_pval = stats::p.adjust(pval, method = "BH"))

```







